Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation

### 2024-04-11 paper/ms.tex
Updates from Overleaf

@@

\title{Inference for Dynamic and Latent Variable Models via Plug-and-Play Automatically Differentiable Particle Filtering}

@@

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{Please provide details of author contributions here.}
\authordeclaration{Please declare any competing interests here.}
%\equalauthors{\textsuperscript{1}A.O.(Author One) contributed equally to this work with A.T. (Author Two) (remove if not applicable).}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: ionides@umich.edu}

@@

\begin{abstract}
  Automatic differentiation has driven recent advances in machine learing, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods.
  This progress has required simultaneous advances in algorithms, software and hardware.  
  Partially observed nonlinear stochastic dynamic systems have proved resistant to these techniques; despite various attempts, widely applicable methods have not yet emerged.
  We present a new approach which is applicable to a general class of models, possesses a theoretical foundation, and is demonstrated to beat current state-of-the-art methods on a challenging scientific benchmark problem.
  Our algorithm is compatible with parallel computation on a graphical processing unit, and enjoys the plug-and-play property that its software implementation requires only a simulator for the scientific dynamic model as input (in addition to data, and a measurement model).

\end{abstract}

@@

\firstpage{7}
% Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting. In this example, there are a total of 11 paragraphs on the first page, counting the first level heading as a paragraph. The value {12} represents the number of the paragraph starting the second page. If a paragraph runs over onto the second page, include a bracket with the paragraph line number starting the second page, followed by the paragraph number in curly brackets, e.g. "\firstpage[4]{11}".

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{M}any approaches to inference in highly nonlinear stochastic dynamical systems assume access to the probability density of next states given the current state.

This is a problem in some critical applications, such as disease modeling, where the models are complex enough that obtaining the density is intractable. However, the particle filter, a popular method for solving the filtering problem in partially-observed dynamical systems, does not require evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm in a feature known as the plug-and-play property.

Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small. Existing methods like the improved iterated filtering algorithm of Ionides et. al. \cite{ionides15} converge quickly to a neighborhood of the MLE, but struggle to optimize the last few units of log-likelihood.

\kevin{TODO: (1) Fix beta smoothness to d(n) at no more than O(n), (2) add story to say that no one spotted that it could have been plug-and-play because of e.g. reasons why they would stumble, (3) }

\kevin{Previous literature: Doucet waymo only applicable to policy plus deterministic dynamics}

We propose a hybrid algorithm that u

Unlike IF2, we explicitly characterize our method's rate of convergence

% IF2 is SGD that iterates through the data sequentially in the 1:N sense, and noisy/mini-batch GD in the 1:J sense. 
% ADPF is batch GD in the 1:N sense, and noisy/mini-batch GD in the 1:J sense.

Algorithmic differentiation potentially facilitates numerical optimization, but currently its use for particle filters is limited. We investigate ways to use algorithmic differentiation of particle filters within the confines of the plug-and-play property, with the goal of enhancing current inference capabilities for general POMP models.

\textbf{The issue:} Particle filters provide convenient approaches to evaluating the log-likelihood function for partially observed Markov process (POMP) models. However, using this evaluator to obtain a maximum likelihood parameter estimate can be challenging -- especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small. Empirically, methods such as the improved iterated filtering algorithm (IF2) from Ionides et. al. \cite{ionides15} rapidly converge to a neighborhood of the optimum but struggle at finding the exact optimum due to Monte Carlo variance, even with an annealing random walk standard deviation.

A potential solution to this could lie in auto-differentiation (AD). This would allow for the use of first and second-order iterative optimization techniques. However, though AD could potentially facilitate numerical optimization, its use for plug-and-play particle filters has so far been limited. This is potentially because particle filtering methods are inherently non-differentiable due to the resampling step that may take place in between iterations.

\subsection*{The importance of the plug-and-play property:} Performing inference in highly nonlinear stochastic dynamical systems is a challenging problem. Although many methods for inference assume access to the density of state transitions, this is often not available, especially in critical applications like epidemiology.

Basic particle filtering algorithms do not require evaluation of the transition density of the latent Markov process, in a feature known as the \textbf{plug-and-play property} \cite{breto09} since it enables an arbitrary model simulator to be plugged into the algorithm. We investigate ways to use algorithmic differentiation of particle filters within the confines of the plug-and-play property, with the goal of enhancing current inference capabilities for general POMP models.

\subsection*{Other potential applications:} This has applications beyond the obvious one of learning model parameters via first or second-order optimization routines. For example, it could be a step towards developing very general Hamiltonian Monte Carlo methods for particle MCMC, as Rosato et. al. \cite{rosato22b} do by using previous work such as \cite{scibior21, poyiadjis11} (and we conjecture that the seed-fixing derivatives of Rosato et. al. are the same as these as an immediate consequence of section \ref{subsec:scibiorrel}) to differentiate the particle filter.

>>>>

\usepackage{tabto}

@@

\title{Accelerated Inference for Partially Observed Stochastic Processes with Automatic Differentiation}

@@

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{K.T. and E.L.I. designed research; K.T. analyzed data; K.T., G.H. and E.L.I. performed research, K.T., G.H., and E.L.I. wrote the manuscript.}
\authordeclaration{The authors declare no competing interests.}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: ionides@umich.edu}

@@

\begin{abstract}
  Automatic differentiation (AD) has driven recent advances in machine learning, including deep neural networks and Hamiltonian Markov Chain Monte Carlo methods. Partially observed nonlinear stochastic dynamical systems have proved resistant to AD techniques due to (1) the requirement for simulation-based inference that does not require access to the system's transition probabilities and (2) the issue that widely used particle filter algorithms yield a likelihood function that is discontinuous as a function of the model parameters. We propose a new representation that embeds existing methods in a theoretical framework that readily permits extension to a new class of algorithms, providing opportunities for optimizing a bias/variance tradeoff. Further, we develop optimization algorithms suited to the Monte Carlo properties of the derivative estimate, including a hybrid algorithm that requires only a differentiable simulator for maximum likelihood estimation. Promising numerical results indicate that a hybrid algorithm that uses AD to refine a coarse solution from iterated filtering can beat current state-of-the-art methods on a challenging scientific benchmark problem.
\end{abstract}

@@

\firstpage{4}
% Use \firstpage to indicate which paragraph and line will start the second page and subsequent formatting. In this example, there are a total of 11 paragraphs on the first page, counting the first level heading as a paragraph. The value {12} represents the number of the paragraph starting the second page. If a paragraph runs over onto the second page, include a bracket with the paragraph line number starting the second page, followed by the paragraph number in curly brackets, e.g. "\firstpage[4]{11}".

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{M}any scientific models involve highly nonlinear stochastic dynamic systems, where there may be significant uncertainty in both the process dynamics and the measurements of the system. Despite the ubiquity of such partially observed stochastic processes with the Markov property (also known as, continuous-state continuous-time hidden Markov models (HMMs), partially-observed Markov processes (POMPs), or partially-observed stochastic nonlinear dynamical systems) in fields as diverse as epidemiology \cite{he10, stocks17}, ecology \cite{knape12}, and finance \cite{kim08, breto14}, performing estimation and inference within this class of models is a difficult problem that faces several challenges.

Many approaches to inference in these scenarios (such as the EM algorithm, the various Kalman filter variants, and MCMC) either struggle with intractable likelihood functions when the models are complex, or assume access to the probability density of next states given the current state. This is a problem in some critical applications, such as disease modeling, where the models are complex enough such that even obtaining the transition density is an intractable problem. However, the particle filter with the bootstrap proposal, a popular method for solving the filtering problem in partially-observed dynamical systems, provides an unbiased estimate of the likelihood \cite{delmoral2004feynman} without requiring evaluation of the transition density of the latent Markov process, enabling an arbitrary model simulator to be plugged into the algorithm.

A series of related algorithms \cite{welch2009abc, wood2010sl, doucet2010pmcmc, ionides08, ionides15} enable parameter estimation within these models. In particular, the improved iterated filtering algorithm (IF2) of Ionides et al. \cite{ionides15} and the particle MCMC algorithm of Andrieu et al. \cite{doucet2010pmcmc} are the only simulation-based full-information maximum likelihood and fully Bayesian methods available in the literature thus far. However, accurate parameter estimation in the presence of significant Monte-Carlo noise in the likelihood estimate remains a significant challenge. Still, maximum likelihood parameter estimation can be challenging, especially when the Monte Carlo variance of the evaluation is high and the number of parameters is not small. For example, though in practice the algorithm of Ionides et al. \cite{ionides15} converges quickly to a neighborhood of the MLE, it often struggles to successfully optimize the last few-units of the log-likelihood. Ionides et al. \cite{Ionides_mcap} attempt to address the third problem, albeit with a computationally-expensive approach.

\subsection{Automatic Differentiation for Particle Filters}

Recent advances in automatic differentiation (AD) for particle filters \cite{blei2018vsmc, jon2018diffpf, corenflos21, scibior2021dpf, doucet2022particlebased} have drawn attention to AD as an alternative tool for inference in partially observed stochastic processes with the particle filter. However, these existing approaches either are not yet compatible with the desire for simulation-based inference, or are computationally expensive. In particular, existing approaches are either asymptotically biased \cite{blei2018vsmc, jon2018diffpf}, have high variance \cite{poyiadjis11, scibior21}, computationally expensive \cite{corenflos21, chen24}, or require transition densities \cite{poyiadjis11, scibior21, doucet2022particlebased, chen24}.

In particular, the work of Scibior and Wood \cite{scibior2021dpf} demonstrates that the estimators derived in Poyiadjis et al. \cite{poyiadjis11} can be attained with AD, albeit with a simple tweak, that they call the stop-gradient trick, to the particle filter that does not modify its likelihood estimates. Though their method is not fully compatible with simulation-based inference and the mathematical motivation behind the stop-gradient trick is not apparent, we show that another simple tweak enables this compatibility and corresponds to a special case of our MOP-$\alpha$ algorithm.

\subsection{Our Contributions}
We develop a new approach to statistical inference via AD for particle filters which addresses these concerns. We construct a new theoretical framework that addresses the seemingly incompatible paradox of differentiating through a Monte-Carlo algorithm with discontinuous resampling.

The key observation is that one can turn this into a problem of differentiating through a series of smooth measurement density ratios and (through the reparametrization trick) a differentiable simulator. This is done through the construction of a suitably reweighted family of particle filters that we call \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$), for some $\alpha \in [0,1]$ that optimizes a bias-variance tradeoff. This family of algorithms encompasses the special case of the biased gradient estimator of \cite{blei2018vsmc} when $\alpha=0$, and the high-variance gradient estimate from \cite{poyiadjis11, scibior21} when $\alpha=1$, and does not require access to the transition densities -- only requiring access to a differentiable simulator.

We show that these off-policy particle filters successfully target the posterior, recover desirable gradient estimates previously encountered in the literature, and possess other desirable theoretical properties. We also derive a linear convergence rate for gradient methods on the particle filter in the presence of strong convexity, providing support for numerical optimization with AD for the particle filter.

In practice, we propose a hybrid algorithm that we call \textbf{Iterated Filtering with Automatic Differentiation (IFAD)} that warm-starts gradient ascent (or a similar iterative first or second-order gradient method) with a coarse solution obtained from a few rounds of iterated filtering. Promising numerical results indicate that IFAD beats IF2 (and by the numerical results of \cite{ionides15}, also IF1, the Liu-West filter \cite{liuwest01}, and particle MCMC \cite{doucet2010pmcmc}) on a challenging scientific benchmark problem, the Dhaka cholera model of \cite{king08}.

\section{Problem Setup}

Consider an unobserved Markov process $(X_t)_{t \geq t_0}$, with observations $Y_1,...,Y_N$ realized at values $y_1^*,...,y_N^*$ at timesteps $t_1,..., t_N$. The process is parameterized by an unknown parameter $\theta \in \Theta \subseteq \R^p$, where the states $X_t$ take values in the state space $\gX \subseteq \R^d$, the observations $Y_n$ take values in $\gY,$ and we write $X_n := X_{t_n}$. By a similar decomposition to that in \cite{doucet2009tutorial}, we find that the joint density of $X_{0:N}, Y_{1:N}$ can be factored as
\begin{align}
    &f_{X_{0: N}, Y_{1: N}}\left(x_{0: N}, y_{1: N} ; \theta\right)\\
    &=f_{X_0}\left(x_0 ; \theta\right) \prod_{n=1}^N f_{X_n \mid X_{n-1}}\left(x_n \mid x_{n-1} ; \theta\right) f_{Y_n \mid X_n}\left(y_n \mid x_n ; \theta\right).
\end{align}

We call $f_{X_n|X_{n-1}}\left(x_{n} \mid x_{n-1}; \theta\right)$ the process model, writing $\texttt{process}\left(x_n; \theta\right)$ for the simulator corresponding to the above process model, $f_{Y_n|X_n}\left(y_n \mid x_n; \theta\right)$ the measurement model, and again write $y_n^*$ for the actual values of the observations that were observed. As a matter of terminology, we call $f_{X_{1:N}|Y_{1:N}}(x_{1_N}|y_{1:N}^*; \theta)$ the posterior distribution of states, and $f_{X_{n}|Y_{1:n}}(x_n|y_{1:n}^*; \theta)$ the filtering distribution at time $t_n$.

The above are all defined on a single probability space $(\Omega, \Sigma, \prob),$ where $\omega \in \Omega$ denotes an element of the sample space that can be thought of as a random seed. For notational simplicity, we omit the dependence of the likelihood and other suitable calculations on $X_0.$ 
In the rest of this paper, we will identify the seed with an element of the sample space. We do so because the seed determines the entire sequence of pseudo-random numbers as generated by a computer, to maintain an algorithmic and practical interpretation of our theory. When two algorithms share the same seed, this corresponds to the method of common random numbers.

\section{Off-Policy Particle Filters}

As mentioned earlier, instead of directly differentiating through a particle filter, one can instead differentiate through the simulator and a series of measurement density ratios. We construct an algorithm called \textbf{Measurement Off-Policy-$\alpha$} (MOP-$\alpha$) in Algorithm \ref{alg:mop} that targets the posterior and likelihood, and whose gradient yields the estimators of \cite{poyiadjis11, scibior21} and \cite{blei2018vsmc} as special cases.

\subsection{Algorithm Outline} 
The broad idea is to run the particle filter twice under the same seed. We first run the particle filter under $\phi$, and then obtain a coupled set of particles by running a bootstrap filter with the process model at $\theta$ but the resampling indices from the previous run at $\phi$. We can then reweight the conditional likelihoods by a correction factor accumulated over time of the ratio of the measurement density under $\theta$ and the measurement density under $\phi$ to account for the ("off-policy") resampling under $\phi$. That is, when we write $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$ as shorthand for the measurement density and estimate the conditional likelihood under $\phi$ in the usual way with $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$, we can estimate the conditional likelihood under $\theta$ as either
\begin{equation}
     \label{eq:mop-conditional-likelihood}
     L_n^{B,\theta,\alpha} = \frac{\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}} \text{ or } L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}},
\end{equation}
where each particle is recursively weighted according to 
\begin{equation}
    \label{eq:weighting-scheme}
    w_{n,j}^{P,\theta} = (w_{n-1,j}^{F,\theta})^\alpha,  w^{F,\theta}_{n,j} = w^{P,\theta}_{n,k_j} { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}, w^{F,\theta}_{0,j}= 1,
\end{equation}
and $k_j$ are the indices from resampling according to the measurement density at $\phi$, $k_j \sim \text{Categorical}(g^{\phi}_{n,1},...,g^{\phi}_{n,J})$. The before-resampling conditional likelihood estimate $L_n^{B,\theta,\alpha}$ should be used in general, as it has slightly lower variance than the after-resampling estimate $L_n^{A,\theta,\alpha} = L_n^\phi\cdot \frac{\sum_{j=1}^J w^{F,\theta}_{n,j}}{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$, but the latter is useful in deriving properties of the MOP-$\alpha$ gradient estimate such as that in Theorem \ref{thm:mop-functional-forms}.

This yields a suitably reweighted particle filter that targets the posterior under $\theta$ and produces a strongly consistent likelihood estimate for the likelihood at $\theta$. Taking the gradient with respect to $\theta$ of the estimate for the log-likelihood at $\theta$ then yields an estimate for the score at $\theta$.

\begin{algorithm}[H]
	\caption{Measurement Off-Policy-$\alpha$}
    \label{alg:mop}
	     \textbf{Input:} Number of particles $J$, timesteps $N$, measurement model $f_{Y_n|X_n}(y_n^*|x_n, \theta)$, simulator $\process(x_{n+1}|x_n, \theta)$, evaluation parameter $\theta$, behavior parameter $\phi$, seed $\omega$. \newline
        \textbf{If $\theta \neq \phi$:}
            Run a particle filter once with parameters $\phi$ to obtain $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$, with seed $\omega$. \newline
        \textbf{Else:} Set $X_{n,j}^{P,\phi}, X_{n,j}^{F,\phi}$ to be copies of $X_{n,j}^{P,\theta}, X_{n,j}^{P,\theta}$ in the rest of the algorithm. \newline
		\textbf{Initialize } filter particles ${X}_{0,j}^{F,\theta}\sim {f}_{{X}_{0}}\left(\cdot\giventh{\theta}\right)$, relative weights $w^{F,\theta}_{0,j}= 1$ for $j$ in $\seq{1}{J}$. Fix $\omega.$ \newline
		\textbf{For} $n=1,...,N$: \newline
            \hspace*{4mm} Accumulate discounted weights, $w_{n,j}^{P,\theta} = \big(w_{n-1,j}^{F,\theta}\big)^\alpha$.\newline
            \hspace*{4mm} Simulate process model,
            ${X}_{n,j}^{P,\theta}\sim {f}_{{X}_{n}|{X}_{n-1}}\big(\cdot|{X}_{n-1,j}^{F};{\theta}\big)$. \newline
            \hspace*{4mm} Measurement density,
            $g^{\theta}_{n,j}={f}_{{Y}_{n}|{X}_{n}}(y_{n}^{*}|{X}_{n,j}^{P,\theta}\giventh{\theta})$. \newline
            \hspace*{4mm} Compute $L_n^{B,\theta,\alpha} ={\sum_{j=1}^Jg^\theta_{n,j} w^{P,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$. \newline
            \hspace*{4mm} Conditional likelihood under $\phi$,
            $L_n^{\phi} = \frac{1}{J}\sum_{m=1}^{J}g^{\phi}_{n,m}$.\newline
            \hspace*{4mm} Select indices $k_{1:J}$ with $\prob\big(k_{j}=m\big) \propto g^{\phi}_{n,m}$. \newline
            \hspace*{4mm} Resample particles ${X}_{n,j}^{F,\phi}={X}_{n,k_{j}}^{P,\phi}$. \newline
            \hspace*{4mm} Resample corrected weights
            $w_{n,j}^{F,\theta}= w^{P,\theta}_{n,k_j}  { g^{\theta}_{n,k_j}}/{ g^{\phi}_{n,k_j}}$.\newline
            \hspace*{4mm} Compute $ L_n^{A,\theta,\alpha} = L_n^\phi\cdot {\sum_{j=1}^J w^{F,\theta}_{n,j}}/{\sum_{j=1}^J  w^{P,\theta}_{n,j}}$.\newline
		\textbf{Return:} likelihood estimate $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\hat{\lik}(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, filtering distribution $\{(X_{N,j}^{F, \theta}, w^{F,\theta}_{N,j})\}_{j=1}^J.$
\end{algorithm}

\subsection{Discounting Weights for a Bias-Variance Tradeoff}

As mentioned earlier, the correction factors accumulate over time. If $\theta$ and $\phi$ coincide, one can discount the correction factors from previous timesteps by some $\alpha \in [0,1]$, as seen in Equation \ref{eq:weighting-scheme}, while still targeting the posterior and likelihood at $\theta$. We do this in order to optimize a bias-variance tradeoff for the score estimate yielded by differentiating through the log-likelihood estimate provided by MOP-$\alpha$. When $\alpha=1$, MOP-$1$ maintains the memory of each particle's ancestral trajectory, recovering the consistent but high-variance gradient estimate from \cite{poyiadjis11, scibior21}. When $\alpha=0$, MOP-$0$ considers only single-step transition dynamics, recovering the low-variance but asymptotically biased gradient estimator of \cite{blei2018vsmc}. $\alpha$ therefore controls an exponentially-weighted moving average, determining how much "memory" the resulting gradient estimate has over ancestral trajectories. This is illustrated in Figure \ref{fig:biasvar}, in the context of the Dhaka cholera model of \cite{king08}.

\begin{figure}[ht!]
    \centering
    \includegraphics[width=\textwidth/3]{imgs/095/biasvar.png}
    \caption{Illustration of the bias-variance tradeoff induced by the discounting hyperparameter $\alpha$, on the Dhaka cholera model of \cite{king08}. We display the MSE of score estimates for the trend in transmission, evaluated at the MLE.}
    \label{fig:biasvar}
\end{figure}

\section{Theoretical Analysis of MOP-$\alpha$}

Here, we will show that MOP-$\alpha$ targets the posterior and likelihood, yields the estimators of \cite{poyiadjis11, scibior21, blei2018vsmc} as special cases when differentiated, and characterize rates for its bias and variance under different choices of $\alpha$. To do so, we require the following assumptions:

\begin{enumerate}[label=(A\arabic*),itemsep=-1.2ex] 
    \item \textbf{Continuity of the Likelihood.} $\ell(\theta)$ has more than two continuous derivatives in a neighborhood $\left\{\theta: \ell(\theta)>\lambda_1\right\}$ for some $\lambda_1<\sup _{\varphi} \ell(\varphi)$. \label{assump:conti-lik}
    \item \textbf{Bounded Process Model.} There exist $\underbar{M}, \bar{M}$ such that $0 < \underbar{M} \leq f_{X_n|X_{n-1}}(x_n | x_{n-1};\theta) \leq \bar{M} < \infty$. \label{assump:bounded-process}
    \item \textbf{Bounded Measurement Model.} There exist $\underbar{G}, \bar{G}$ such that $0<\underbar{G} \leq f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right) \leq \bar{G}<\infty$ and $G'(\theta)$ so $||\nabla_\theta \log f_{Y_n \mid X_n}\left(y_n^* \mid x_n; \theta\right)||_\infty \leq G'(\theta)< \infty$. \label{assump:bounded-measurement}
    \item \textbf{Bounded Gradient Estimates.} There are functions $G(\theta), H(\theta): \Theta \to [0,\infty)$ uniformly bounded by $G', H'$, so the MOP-$\alpha$ gradient and Hessian estimates at $\theta=\phi$ are almost surely bounded by $G(\theta)$ and $H(\theta)$ for all $\alpha$. \label{assump:local-bounded-derivative}
    \item \textbf{Differentiability of Density Ratios and Simulator.} The measurement density $f_{Y_n|X_n}(y_n^*|x; \theta)$ and simulator have more than two continuous derivatives in $\theta$. \label{assump:diff-meas-and-sim}
\end{enumerate}

The fact that the likelihood estimate yielded by MOP-$\alpha$ has more than two continuous derivatives in $\theta$ follows the construction of the likelihood estimate in equations \ref{eq:mop-conditional-likelihood} and \ref{eq:weighting-scheme}, as well as Assumptions \ref{assump:bounded-measurement} and \ref{assump:diff-meas-and-sim}.

\subsection{MOP-$\alpha$ Targets the Posterior}

We show here that MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood. To do so, we state the definition of targeting and present three lemmas, whose proof can be found in the supplementary material.
\begin{defn}[Targeting]
    \label{defn:targeting}
    A random vector $(X, w)$ drawn from a distribution $g$ \textbf{targets} the distribution $\pi$ if for any measurable and bounded functional $h$,
    \begin{equation}
        E_g\{h(X) \cdot w\}=E_\pi\{h(X)\}
    \end{equation}  
        A set of particles $(X^J_j, w^J_j), j=1,2, \ldots,J$, \textbf{targets} $\pi$ if for any measurable and bounded functional $h$, as $J\to\infty$
    \begin{equation}
        \frac{\sum_{j=1}^J h(X^J_j) w^J_j}{\sum_{j=1}^J w^J_j} \stackrel{a.s.}{\to} E_\pi(h(X)).
    \end{equation}
\end{defn}

\begin{lem}[Change of Particle Measure]
    \label{lem:change-measure-proper-weights}
    Assume $\{(\tilde X_j^J,U_j^J)\}_{j=1}^J$ targets $f_X$, and that $\{(Y_j^J,V_j^J)\}_{j=1}^J$ is a multinomial sample drawn from $\{(\tilde X_j^J,U_j^J)\}$ where $(\tilde X_j^J,U_j^J)$ is drawn w.p. $\pi^J_j$. Write, where $a$ is the ancestor function,
    $(Y_j^J,V_j^J) = \big(\tilde X^J_{a(j)},U^J_{a(j)}/\pi^J_{a(j)}\big)
    $. If the importance sampling weights $U_j/\pi_j$ are bounded, then $\{(Y^J_j,V^J_j)\}_{j=1}^J$ targets $f_X$.
\end{lem}
\begin{lem}[Particle Marginals]
    \label{lem:marginal-proper-weights}
    Suppose that $\{(\tilde X_j^J,U_j^J)\}_{j=1}^J$ targets $f_X$. Also suppose that $\tilde Z_j^J \sim f_{Z|X}(\cdot | \tilde X_j^J)$, where $f_{Z|X}$ is a conditional probability density function corresponding to a joint density $f_{X,Z}$ with marginal densities $f_X$ and $f_Z$. Then, if the $U_j^J$ are bounded, $\{(\tilde Z_j^J,U_j^J)\}_{j=1}^J$ targets $f_Z$.
\end{lem}
\begin{lem}[Particle Posteriors]
    \label{lem:posterior-proper-weights}
    Suppose that $\{(X_j^J,u_j^J)\}_{j=1}^J$ targets $f_X$, and write $(X^{\prime J}_j,u^{\prime J}_j) = \big(X_j^J,u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$. Then, if $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big)$ and $u_j^J \cdot f_{Z|X}(z^*|X_j^J)\big) / f_Z(z^*)$ are bounded, $\{(X^{\prime J}_j,u^{\prime J}_j)\}_{j=1}^J$ targets $f_{X|Z}(\cdot | z^*)$.
\end{lem}

From these lemmas, we can derive the following result.

\begin{thm}[MOP-$\alpha$ Targets the Posterior and Likelihood]
    \label{thm:mop-targeting}
    When $\alpha=1$ or $\phi=\theta$, MOP-$\alpha$ targets the posterior and is strongly consistent for the likelihood. That is, for $\pi=f_{X_{1:N}|Y_{1:N} ; \theta}$ and any measurable and bounded functional $h$ and for $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$ or $\prod_{n=1}^N L_n^{B,\theta,\alpha}$, it holds that
    \begin{equation}
        \frac{\sum_{j=1}^J h(x_{n,j}^{P, \theta}) w_{n,j}^{P,\theta}}{\sum_{j=1}^J w_{n,j}^{P,\theta}} \stackrel{a.s.}{\to} E_\pi h, \;\; \hat\lik(\theta)  \stackrel{a.s.}{\to} \lik(\theta).
    \end{equation}
\end{thm}

\begin{proof}
    When $\theta=\phi$, regardless of the value of $\alpha$, the ratio ${g_{n,j}^\theta}/{g_{n,j}^\phi}=1,$ and this reduces to the vanilla particle filter estimate.

    When $\alpha=1$, and $\theta\neq\phi,$ the proof is as follows. Recursively applying Lemmas \ref{lem:change-measure-proper-weights}, \ref{lem:marginal-proper-weights}, and \ref{lem:posterior-proper-weights}, we obtain that 
    %to step~\ref{mop:step1}, Lemma~2 step~ {mop:weight:update} and Lemma~3 to step~\ref{mop:step2} we obtain that
    the MOP-1 filter targets the posterior.
    Specifically, suppose inductively that $\big\{\big(X^{F,\theta}_{n-1,j},w^{F,\theta}_{n-1,j}\big)\big\}$ is properly weighted for $f_{X_{n-1}|Y_{1:n-1}}(x_{n-1}|y^*_{1:n-1};\theta)$.
    Then, Lemma \ref{lem:marginal-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big)\big\}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.
    Lemma \ref{lem:posterior-proper-weights} tells us that $\big\{\big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j} g^\theta_{n,j} \big)\big\}$ therefore targets  $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.
    Lemma \ref{lem:change-measure-proper-weights} guarantees that the resampling rule, given by 
    \begin{equation}
        \big(X^{F,\theta}_{n,j},w^{F,\theta}_{n,j}\big) = \big(X^{P,\theta}_{n,a(j)}, w^{P,\theta}_{n,a(j)} g^\theta_{n,a(j)}\big/ g^\phi_{n,a(j)}\big),
    \end{equation}
    with resampling weights proportional to $g^\phi_{n,j}$, therefore also targets $f_{X_{n}|Y_{1:n}}(x_{n}|y^*_{1:n};\theta)$.

    For the likelihood, if the likelihood is estimated with the before-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{B,\theta,\alpha}$, the strong consistency is a direct consequence of our earlier result that $\{ \big(X^{P,\theta}_{n,j},w^{P,\theta}_{n,j}\big) \}$ targets $f_{X_{n}|Y_{1:n-1}}(x_{n}|y^*_{1:n-1};\theta)$.

    Whereas if we use the after-resampling conditional likelihoods $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A,\theta,\alpha}$, the numerator of $L_n^{B,\theta,\alpha}$ is
    \begin{equation}
        \sum_{j=1}^J g_{n,j}^\theta w_{n, j}^{P, \theta} = L^\phi_n \sum_{j=1}^J \left[ \frac{g^\theta_{n,j}}{g^\phi_{n,j}} w^{P,\theta}_{n,j}\right] \frac{g^\phi_{n,j}}{L_n^\phi}
    \end{equation}
    Using Lemma \ref{lem:change-measure-proper-weights}, we resample the corrected weights $w^{P,\theta}_{n,j}{g^\theta_{n,j}}/{g^\phi_{n,j}} $ according to probabilities ${g^\phi_{n,j}}/{L_n^\phi}$ to see this is properly estimated by $L^\phi_n \sum_{j=1}^J w^{F,\theta}_{n,j},$
from which we obtain the estimate $L_n^{B,\theta,\alpha}$. The strong consistency then follows as the likelihood is the product of the conditional likelihoods.
\end{proof}

\subsection{MOP-$\alpha$ Encompasses the Estimators of \cite{poyiadjis11, scibior21, blei2018vsmc}}

\cite{scibior21} show that the estimate of \cite{blei2018vsmc} is the gradient of a vanilla particle filter when resampling terms are dropped, and also recover the estimate of \cite{poyiadjis11} through applying a "stop-gradient trick" to the particle filter. It turns out that both of these, when applied on the bootstrap filter, correspond to special cases of MOP-$\alpha$.

\begin{thm}[MOP-$0$ and MOP-$1$ Functional Forms]
    \label{thm:mop-functional-forms}
    Write $\nabla_\theta \hat\ell^\alpha(\theta)$ for the gradient estimate yielded by MOP-$\alpha$ when $\theta=\phi$ and we use the after-resampling conditional likelihood estimate so that $\hat\lik(\theta) = \prod_{n=1}^N L_n^{A, \theta, \alpha}$. Then, when $\alpha=0$,
    \begin{equation}
        \nabla_\theta \hat\ell^0(\theta) 
        = \frac{1}{J} \sum_{n=1}^N \sum_{j=1}^J \nabla_\theta \log\left(f_{Y_n|X_{n}}(y_n^*|x_{n,j}^{F, \theta}; \theta)\right),
    \end{equation}
    yielding the gradient estimator of \cite{blei2018vsmc} with the bootstrap filter. When $\alpha=1$,
    \begin{equation}
        \nabla_\theta \hat{\ell}^1(\theta) 
        = \frac{1}{J}\sum_{j=1}^J \nabla_\theta \log f_{Y_{1:N}|X_{1:N}}\left(y_{1:N}^* | x_{1:n,j}^{A, F,\theta}\right),
    \end{equation}
    yielding the gradient estimator of \cite{poyiadjis11, scibior21} applied to the bootstrap filter.
\end{thm}

The proof follows from a useful decomposition of the after-resampling conditional likelihood estimate $L_n^{A,\theta,\alpha}$ that yields a telescoping product in the MOP-$1$ case. Repeated applications of the log-derivative identity that $\nabla_x \log(f(x)) = (\nabla_x f(x))/f(x)$, and noting that when $\theta=\phi$, that $w_{n,j}^{P,\theta}$ evaluates to $1$, yield the result. As such, we defer the proof to the supplementary material.

This further illustrates the role of $\alpha$ as a paramteter that dictates the "memory" of the gradient estimate. As \cite{scibior2021dpf} remark, the MOP-$0$ estimator of \cite{blei2018vsmc} only considers single-step quantities, and is "memoryless" beyond a single step. This is in contrast to the case when $\alpha=1,$ as that estimate of \cite{poyiadjis11} only considers the surviving particles at time $N$, fully tracking dependencies over time.

\subsection{MOP-$1$ Is Consistent for the Score}
